How to build an homogeneous language for heterogeneous platforms

### 2013-06-11 papers/gpce2013/gpce2013.tex

@@

The last approach, shown in figure \ref{langs} (d), can be seen as a middle-ground between the two previous
approaches: DSLs are embedded in a host language but use a code generation process. (benefits and limitations of
this approach)

Lightweight Modular Staging~\cite{Rompf12_LMSThesis} (LMS) is a framework for defining deeply embedded DSLs in Scala
using staging~\cite{Jorring1986_Staging}. It has been used to define high-performance DSLs for parallel
computing~\cite{Brown11_Parallel}. (Todo. Mention \code{Rep[_]})

Code generation consists in traversing the statement nodes produced by the program evaluation according to their
dependencies and to emit the code corresponding to each statement. LMS already sorts the statements graph so DSL
authors just need to say how to emit code for each statement node of their DSL.

@@

\begin{figure}
\begin{lstlisting}[label=selector-impl,caption=Selectors optimization]
def find(receiver: Exp[Selector],
         selector: Exp[String]) =
  getConstIdCss(selector) match {
    case Some(id) if receiver == document =>
      DocumentGetElementById(Const(id))
    case _ =>
      SelectorFind(receiver, selector)
  }
\end{lstlisting}
\end{figure}

@@

We also have written several implementations of a complete application using different approaches for the client and
server sides, and compared the amount of code written, the runtime performances and the ability to modularize the
code and to maintain it.

@@

Figure \ref{micro-benchmark} shows the micro-benchmark results. We measured the runtime performance, the number of
lines of code (as an indication of the abstraction level), and finally the ratio of runtime performance and lines of
code. \emph{js-scala} is between 3 to 100 times faster than other approaches on our benchmark. Combined with that
fact that it provides high-level abstractions, it has a performance / lines of code ratio more than 4 times higher
than other approaches.

@@

The application was initially written using jQuery. We rewrote it using several technologies for the client-side
part: plain JavaScript (without third-party library), js-scala, GWT and HaXe.

@@

Our goal is to evalute the level of abstraction provided by each solution and their performances. We took the
number of lines of code as a measure of the level of abstraction. We also measured the size of the data sent to the
client-side. We also measured the ability to share code between client and server sides.

Rather than doing a micro-benchmark focusing on just one abstraction, we performed a global benchmark that is more
likely to reflect the real performances of the application: we simulated user actions on a Web page (2000 clicks on
buttons, triggering a dynamic update of the page and involving the use of the Option monad, the selectors API and
the HTML fragment definition API) and measured the time it took complete them. We only measured the execution time of
client-side code execution. The tests were run on a DELL Latitude E6430 laptop with 8 GB of RAM, on the Google
Chrome v27 Web browser. Figure \ref{benchmark} shows the benchmark results.

@@

>>>>

\subsection{Language Engineering Processes}

@@

The last approach, shown in figure \ref{langs} (d), can be seen as a middle-ground between the two previous
approaches: DSLs are embedded in a host language but use a code generation process. The embedding makes it easy to
combine several DSLs and makes the language implementation easier because the host language infrastructure is
reused. On the other hand, the syntax of the DSLs is limited by the syntax flexibility of the host language. The code
generation process is specific to each DSL and gives the opportunity to perform domain-specific optimizations.

(Summary explaining why we choose the deep embedded DSLs approach)

\subsection{Lightweight Modular Staging}

Lightweight Modular Staging~\cite{Rompf12_LMSThesis, Rompf12_LMS} (LMS) is a framework for defining deeply embedded
DSLs in Scala. It has been used to define high-performance DSLs for parallel computing~\cite{Brown11_Parallel} and to
define JavaScript as an embedded DSL in Scala~\cite{Kossakowski12_JsDESL}.

LMS is based on staging~\cite{Jorring1986_Staging}: a program using LMS is a regular Scala program that evaluates
to an intermediate representation (IR) of a final program. A code generator can traverse this IR to produce the final
program code. Expressions evaluated in the initial program and those evaluated in the final program are distinguished
by their type: a \code{Rep[Int]} value in the initial program is a staged computation that generates code evaluating
to an \code{Int} value in the final program. An \code{Int} computation in the initial program is evaluated during the
initial program evaluation and becomes a constant in the final program.

Defining a DSL with LMS consists in the following steps:

\begin{itemize}
 \item writing a Scala module providing the DSL vocabulary as an abstract API,
 \item implementing the API in terms of IR nodes,
 \item defining a code generator visiting IR nodes and generating the corresponding code.
\end{itemize}

@@

\begin{figure}
\begin{lstlisting}[label=selector-impl,caption=Selectors optimization]
def find(receiver: Rep[Selector],
         selector: Rep[String]) =
  getConstIdCss(selector) match {
    case Some(id) if receiver == document =>
      DocumentGetElementById(Const(id))
    case _ =>
      SelectorFind(receiver, selector)
  }
\end{lstlisting}
\end{figure}

@@

We also have written several implementations of a complete application using different approaches for the client and
server sides, and compared the amount of code written, the runtime performances and the ability to modularize the
code.

@@

Figure \ref{micro-benchmark} shows the micro-benchmark results. We measured the runtime performance, the number of
lines of code (as an indication of the abstraction level), and finally the ratio of runtime performance and lines of
code. \emph{js-scala} is between 3 to 100 times faster than other approaches on our benchmark. Combined with the
fact that it provides high-level abstractions, it has a performance / lines of code ratio more than 4 times higher
than other approaches.

@@

The application was initially written using jQuery. We rewrote it using several technologies for the client-side
part: vanilla JavaScript (low-level code without third-party library), js-scala, GWT and HaXe. In each case we tried
to write the application in an idiomatic way.

@@

Our goal is to evaluate the level of abstraction provided by each solution and their performances. We took the
number of lines of code as a measure of the level of abstraction. We also measured the ability to share code between
client and server sides.

We simulated user actions on a Web page (2000 clicks on buttons, triggering a dynamic update of the page and
involving the use of the Option monad, the selectors API and the HTML fragment definition API) and measured the time
it took complete them. We only measured the execution time of client-side code execution. The tests were run on a
DELL Latitude E6430 laptop with 8 GB of RAM, on the Google Chrome v27 Web browser. Figure \ref{benchmark} shows the
benchmark results.

@@

The runtime performances of the vanilla JavaScript, HaXe and \emph{js-scala} versions are similar (though the
\emph{js-scala} version is slightly slower by 6\%). It is worth noting that the vanilla JavaScript and the HaXe
versions use low-level code compared to \emph{js-scala}, as shown in the second part of the figure (lines of code):
the \emph{js-scala} version needs only 74 lines of code while the vanilla JavaScript version needs 116 lines of code
(57\% bigger) and the HaXe version needs 148 lines of code (100\% bigger). The jQuery JavaScript version, of which
code is high-level (54 lines of code, 27\% less than \emph{js-scala}) runs 10 times slower than the \emph{js-scala}
version.

The last part of the figure compares the runtime performance / lines of code ratio. \emph{js-scala} shows the best
score, being 1.48 times better than the vanilla JavaScript version, 1.88 times better than the HaXe version,
3.77 times better than the GWT version and 7.82 times better than the jQuery JavaScript version.
